Validation of Visual Statistical Inference, with Application to Linear Models

### 2012-03-09 paper_jasa.tex
parametric vs non-parametric tests

\begin{dfn} \label{dfn:test}
A \textit{visual test} $T_\theta$ is defined as the grammar \citep{wilkinson:1999} of the plot used in the lineup. The grammar consists of type and specification of aesthetics necessary for complete reproducibility.
\end{dfn}
}
Next, we will develop power theoretically and then relate it to the empirical results.

\begin{dfn} \label{dfn:power}
\red{HH: I don't think power should be defined in terms of $\theta$ Usually power is considered a property of the test.  In our case that's the specifics of the plot used for the lineup. We might need an additional definition for that - maybe we can phrase that within the framework of the grammar of graphics? - that's what I've tried above}

For a lineup of $m$ plots the power of \blue{a visual test}
%$\theta$
 is defined as 
    \begin{equation*}
      \text{Power}(T_\theta)= 
        \begin{cases} 
              \text{Type-I error}=\frac1m & \text{if $\theta \in \Theta_0$,} \\
              Pr(\text{Reject } H_0) &\text{if $\theta \in \Theta^c_0$.}
        \end{cases}
    \end{equation*}
\end{dfn}
\red{this is a non-standard definition of power - usually power is just the prob to reject under $H_1$. could we back the non-standard definition up with some literature?}

@@

\blue{Let $D$ be a measure for the effect contained in a data set. We will assume that $D$ follow a distribution $F_D$. Let $p_D$ be the $p$-value associated with the data shown in the test statistic.

Similarly, we will assume that $p_{0,i}$, $i=1, ..., m-1$ are the $p$-values associated with data corresponding to the $m-1$ null plots. Since this data is generated consistently with the null hypothesis, the $p$-values follow a standard Uniform distribution, $p_{i,0} \sim U[0,1], i= 1, ..., m-1$. }

\red{move power consideration up:}

@@

Under $H_1$ we therefore get an estimate for the strength of the signal in the test statistic based on a lineup evaluation as 
\[
\widehat{E[p_D]} = 1- \left(\frac{n_m}{K}\right)^{1/(m-1)}.
\]

\red{We need comment on how this ties in to the rest of the paper: this is a theoretical consideration based on the assumption that observers are able to spot the plot with the smallest/a small $p$-value.}

@@

% Di: This is specific for the regression model
%Consider the hypothesis test $H_0: \beta_k=0$ against $H_1: \beta_k \ne 0$ in model \ref{multi}.
Now consider estimating the power of the visual test. 
Suppose $F_{|t|}(.)$ is the distribution function of an absolute value of $t$. For the regression slope parameter $\beta$ suppose the associated test statistic in classical test be $t_{obs}$ with $p$-value $p_B$. By definition we have

\red{HH: the following paragraph should probably move up, but I'm not sure how to change it appropriately. Mahbub, could you take care of that?}
Now suppose $F_{|t|}(.;\delta)$ denotes the distribution function of an absolute value of $t$ distribution with non-centrality parameter $\delta$. Thus the distribution function of $p-value$ be  
\begin{eqnarray}\label{dist_p}
F{p_B}(p) &=& Pr(p_B \le p)=1-Pr(1-p_B \le 1-p) \nonumber \\
  &=& 1-Pr(F_{|t|}^{-1}(1-p_B) \le F_{|t|}^{-1}(1-p) ) \nonumber \\
  &=& 1-Pr(|t_{obs}| \le F_{|t|}^{-1}(1-p)) \nonumber \\
  &=&\left\{ \begin{array}{ll}
          1-F_{|t|}( F_{|t|}^{-1}(1-p))=p &\mbox{ ; under $H_0$} \\
          1-F_{|t|}( F_{|t|}^{-1}(1-p); \delta) &\mbox{ ; under $H_1$} 
       \end{array} \right.     
\end{eqnarray}

Thus the density of $p_B$ is Uniform(0,1)  under $H_0$. As noted by \cite{Ruppert:2007}, under $H_1$ the density $$f_{P_B}(p_B; \delta)= \frac{f_{|t|}(F_{|t|}^{-1}(1-p_B);\delta)}{f_{|t|}(F_{|t|}^{-1}(1-p_B))}$$ derived from equation \ref{dist_p}  is a right skewed distribution.  
%In a lineup plot we simulate $m-1$ residual data sets from null model where each of these $m-1$ null data sets produces a corresponding $p$-value $p_{0,i}$ and $p_{0,i} \sim \text{Uniform}(0,1)$ for $i = 1, ..., m-1$. Suppose  $p_0=\min_i(p_{0,i})$. Thus $p_0 \sim \text{Beta}(1,m-1)$.

>>>>

\begin{dfn} \label{dfn:test}
A \textit{visual test} $T_\theta$ is defined as the grammar \citep{wilkinson:1999} of the plot used in the lineup. The grammar consists of type and specification of aesthetics necessary for complete reproducibility.
\end{dfn}
}
%Next, we will develop power theoretically and then relate it to the empirical results.
\blue{We will first theoretically define and  derive the power of  a visual test in the setting of a parametric hypothesis test, i.e. we assume a test  of $H_0: \theta = 0$ vs $H_1: \theta \neq 0$ for some real valued parameter vector $\theta$. In a next step we will then relate the theoretical results to our empirical findings.}

\begin{dfn} \label{dfn:power}
For a lineup of $m$ plots the power of \blue{a visual test} $T$
%$\theta$
 is defined as 
    \begin{equation*}
      \text{Power}_T(\theta)= 
        \begin{cases} 
              \text{Type-I error}=\frac1m & \text{if $\theta \in \Theta_0$,} \\
              Pr(\text{Reject } H_0) &\text{if $\theta \in \Theta^c_0$.}
        \end{cases}
    \end{equation*}
\end{dfn}

@@

\red{HH: the following paragraph needs some adjustment in terms of t and $\theta$ - could you try to take of that, Mahbub?}
Suppose $F_{|t|}(.)$ is the distribution function of an absolute value of $t$. For the regression slope parameter $\beta$ suppose the associated test statistic in classical test be $t_{obs}$ with $p$-value $p_B$. By definition we have

Now suppose $F_{|t|}(.;\delta)$ denotes the distribution function of an absolute value of $t$ distribution with non-centrality parameter $\delta$. Thus the distribution function of $p-value$ be  
\begin{eqnarray}\label{dist_p}
F{p_B}(p) &=& Pr(p_B \le p)=1-Pr(1-p_B \le 1-p) \nonumber \\
  &=& 1-Pr(F_{|t|}^{-1}(1-p_B) \le F_{|t|}^{-1}(1-p) ) \nonumber \\
  &=& 1-Pr(|t_{obs}| \le F_{|t|}^{-1}(1-p)) \nonumber \\
  &=&\left\{ \begin{array}{ll}
          1-F_{|t|}( F_{|t|}^{-1}(1-p))=p &\mbox{ ; under $H_0$} \\
          1-F_{|t|}( F_{|t|}^{-1}(1-p); \delta) &\mbox{ ; under $H_1$} 
       \end{array} \right.     
\end{eqnarray}

Thus the density of $p_B$ is Uniform(0,1)  under $H_0$. As noted by \cite{Ruppert:2007}, under $H_1$ the density $$f_{P_B}(p_B; \delta)= \frac{f_{|t|}(F_{|t|}^{-1}(1-p_B);\delta)}{f_{|t|}(F_{|t|}^{-1}(1-p_B))}$$ derived from equation \ref{dist_p}  is a right skewed distribution.

\blue{Similarly, we will assume that $p_{0,i}$, $i=1, ..., m-1$ are the $p$-values associated with data corresponding to the $m-1$ null plots. Since this data is generated consistently with the null hypothesis, the $p$-values follow a standard Uniform distribution, $p_{i,0} \sim U[0,1], i= 1, ..., m-1$. }

\red{HH: $p_D$ isn't defined yet}

@@

Under $H_1$ we therefore get an estimate for the strength of the signal in the test statistic based on a lineup evaluation as 
\begin{equation}\label{eqn:power_estimate}
\widehat{E[p_D]} = 1- \left(\frac{n_m}{K}\right)^{1/(m-1)}.
\end{equation}

\blue{Note that the right hand side of equation (\ref{eqn:power_estimate}) is independent of the parameter $\theta$, but just based on the lineup evaluation by independent observers. This allows us to compute this value independently of the parametric setting. We therefore define the {\it signal strength} of a  visual test as the following:

\begin{dfn} \label{dfn:signal}
The {\it signal strength} $\sigma_T$ of a visual test is defined as the probability to identify the data plot.
\end{dfn}

In a lineup of size $m$, evaluated by $K$ independent observers,  we can estimate the signal strength of the test as
\[
\sigma_T^{m-1} = \frac{n_m}{K}.
\]

In a parametric setting, signal strength of a test coincides with the power of the test.
}

\red{We need to comment on how this ties in to the rest of the paper: this is a theoretical consideration based on the assumption that observers are able to spot the plot with the smallest/a small $p$-value.}

@@

% Di: This is specific for the regression model
%Consider the hypothesis test $H_0: \beta_k=0$ against $H_1: \beta_k \ne 0$ in model \ref{multi}.
%%Now consider estimating the power of the visual test. 
%%Suppose $F_{|t|}(.)$ is the distribution function of an absolute value of $t$. For the regression slope parameter $\beta$ suppose the associated test statistic in classical test be $t_{obs}$ with $p$-value $p_B$. By definition we have 
%%
%%$$p_B=Pr(|t| \ge t_{obs}| H_0)=1-F_{|t|}(t_{obs}) \Rightarrow |t_{obs}|=F_{|t|}^{-1}(1-p_B)$$
%%
%%\red{HH: the following paragraph should probably move up, but I'm not sure how to change it appropriately. Mahbub, could you take care of that?}
%%Now suppose $F_{|t|}(.;\delta)$ denotes the distribution function of an absolute value of $t$ distribution with non-centrality parameter $\delta$. Thus the distribution function of $p-value$ be  
%%\begin{eqnarray}\label{dist_p}
%%F{p_B}(p) &=& Pr(p_B \le p)=1-Pr(1-p_B \le 1-p) \nonumber \\
%%  &=& 1-Pr(F_{|t|}^{-1}(1-p_B) \le F_{|t|}^{-1}(1-p) ) \nonumber \\
%%  &=& 1-Pr(|t_{obs}| \le F_{|t|}^{-1}(1-p)) \nonumber \\
%%  &=&\left\{ \begin{array}{ll}
%%          1-F_{|t|}( F_{|t|}^{-1}(1-p))=p &\mbox{ ; under $H_0$} \\
%%          1-F_{|t|}( F_{|t|}^{-1}(1-p); \delta) &\mbox{ ; under $H_1$} 
%%       \end{array} \right.     
%%\end{eqnarray}
%%
%%Thus the density of $p_B$ is Uniform(0,1)  under $H_0$. As noted by \cite{Ruppert:2007}, under $H_1$ the density $$f_{P_B}(p_B; \delta)= \frac{f_{|t|}(F_{|t|}^{-1}(1-p_B);\delta)}{f_{|t|}(F_{|t|}^{-1}(1-p_B))}$$ derived from equation \ref{dist_p}  is a right skewed distribution.  
%In a lineup plot we simulate $m-1$ residual data sets from null model where each of these $m-1$ null data sets produces a corresponding $p$-value $p_{0,i}$ and $p_{0,i} \sim \text{Uniform}(0,1)$ for $i = 1, ..., m-1$. Suppose  $p_0=\min_i(p_{0,i})$. Thus $p_0 \sim \text{Beta}(1,m-1)$.
